Understanding the `aghq` package

### 2022-09-30 src/explore_aghq/aghq.Rmd
Improve aghq walkthrough

# Basic

```{r}
set.seed(84343124)
y <- rpois(10, 5)
plot(y)

log_post <- function(eta, y) {
  sum(y) * eta - (length(y) + 1) * exp(eta) - sum(lgamma(y + 1)) + eta
}

fn <- function(x) log_post(x, y)
gr <- function(x) numDeriv::grad(fn, x)
he <- function(x) numDeriv::hessian(fn, x)

ff <- list(fn = fn, gr = gr, he = he)

quad <- aghq::aghq(ff = ff, k = 3, startingvalue = 0)

class(quad)

summary(quad)
plot(quad)

# Containing
str(quad)
# normalized_posterior
quad$normalized_posterior$nodesandweights
quad$normalized_posterior$lognormconst
# marginals
# optresults
quad$optresults$mode
```

# Background

```{r}

# Quadrature

# Single dimensional example
f <- function(x) {
  exp(-(x^2) / 2)  
}

# Grid for GHQ with 3 points
gg <- mvQuad::createNIGrid(1, "GHe", 3)
mvQuad::quadrature(f, gg)
sqrt(2 * pi)

mvQuad::getNodes(gg)
mvQuad::getWeights(gg)

# Two dimensional example
f <- function(x) {
  apply(x, 1, function(y) prod(exp(-(y^2) / 2)))  
}

# Still 3 points (per dimension I'd guess)
gg2 <- mvQuad::createNIGrid(2, "GHe", 3)

mvQuad::getNodes(gg2)
# which is equivalent to
expand.grid(mvQuad::getNodes(gg),mvQuad::getNodes(gg))

mvQuad::getWeights(gg2)
# which is equivalent to
cbind(apply(expand.grid(mvQuad::getWeights(gg),mvQuad::getWeights(gg)),1,prod))

mvQuad::quadrature(f, gg2)
2 * pi

# Gauss Hermite Quadrature
# Showing that the nodes are zeros of Hermite polynomial
Hk <- as.function(mpoly::hermite(3, kind = "he"))
nn <- mvQuad::getNodes(gg)
round(Hk(nn))

>>>>

```{r}
cbpalette <- multi.utils::cbpalette()
```

# Basic usage

Data:

```{r}
set.seed(84343124)
y <- rpois(10, 5)
plot(y)
```

The log-posterior for the model is:

```{r}
log_post <- function(eta, y) {
  sum(y) * eta - (length(y) + 1) * exp(eta) - sum(lgamma(y + 1)) + eta
}
```

Get the first two derivatives too, and put them into a list:

```{r}
fn <- function(x) log_post(x, y)
gr <- function(x) numDeriv::grad(fn, x)
he <- function(x) numDeriv::hessian(fn, x)

ff <- list(fn = fn, gr = gr, he = he)
```

Use the `aghq` package with `k = 3`:

```{r}
quad <- aghq::aghq(ff = ff, k = 3, startingvalue = 0)
```

The result is of class `"aghq"`:

```{r}
class(quad)
```

```{r}
summary(quad)
plot(quad)
```

The `quad` object contains:

```{r}
str(quad)
```

The nodes and weights of the posterior are:

```{r}
quad$normalized_posterior$nodesandweights
```

The logarithm of the normalising constant (should correspond to `logpost - logpost_normalized`):

```{r}
quad$normalized_posterior$lognormconst
```

The mode is as follows, which corresponds to the second node.
For `k = 3`, and other odd numbers, the mode is included as a node.

```{r}
quad$optresults$mode
```

# More detailed background

## One dimensional example

Suppose we have
$$
f(x) = \exp(-\frac{x^2}{2}),
$$
which we would like to integrate over $\mathbb{R}$, i.e. $\int_{-\infty}^\infty f(x) \text{d}x$.

```{r}
f <- function(x) {
  exp(-(x^2) / 2)  
}
```

```{r}
plot <- data.frame(x = seq(from = -10, to = 10, by = 0.1)) %>%
  mutate(
    y = f(x)
  ) %>%
  ggplot(aes(x = x, y = y)) +
    geom_line() +
    theme_minimal()

plot
```

Going to do Gauss-Hermite quadrature with `k = 3` points:

```{r}
gg <- mvQuad::createNIGrid(1, "GHe", 3)
```

The positions of the nodes and their weights are:

```{r}
gg$nodes
gg$weights

#' The scale for weights is a bit silly
plot +
  geom_point(
    data = data.frame(nodes = gg$nodes, weights = gg$weights, y = 0), 
    aes(x = nodes, y = y, size = weights),
    col = cbpalette[2]
  )
```

We can perform the quadrature using `mvQuad` via:

```{r}
mvQuad::quadrature(f, gg)
```

Or more manually via:

```{r}
sum(f(gg$nodes) * gg$weights)

#' Note: you can also use these functions to get the nodes and weights
# mvQuad::getNodes(gg)
# mvQuad::getWeights(gg)
```

This is surprisingly close to the real answer, which is:

```{r}
sqrt(2 * pi)
```

## Two dimensional example

What about in two dimensions?

```{r}
f <- function(x) {
  prod(exp(-(x^2) / 2))
}

df <- expand.grid(
  x1 = seq(-3, 3, by = 0.1),
  x2 = seq(-3, 3, by = 0.1)
)

plot2 <- df %>%
  rowwise() %>%
  mutate(
    y = f(c(x1, x2))
  ) %>%
  ggplot(aes(x = x1, y = x2, fill = y)) +
    geom_tile() +
    theme_minimal()

plot2
```

```{r}
gg2 <- mvQuad::createNIGrid(2, "GHe", 3)
gg2
```

The nodes and weights are:

```{r}
mvQuad::getNodes(gg2)
mvQuad::getWeights(gg2)
```

Equivalent to:

```{r}
expand.grid(mvQuad::getNodes(gg), mvQuad::getNodes(gg))
cbind(apply(expand.grid(mvQuad::getWeights(gg), mvQuad::getWeights(gg)), 1, prod))
```

```{r}
plot2 +
  geom_point(
    data = mvQuad::getNodes(gg2) %>%
            as.data.frame() %>%
            mutate(weights = mvQuad::getWeights(gg2)), 
    aes(x = V1, y = V2, size = weights, fill = 1),
    col = cbpalette[2]
  )
```

Do the quadrature using:

```{r}
#' Need f to be defined like this for some reason
f <- function(x) {
  apply(x, 1, function(y) prod(exp(-(y^2) / 2)))  
}

mvQuad::quadrature(f, gg2)
```

As compare with the true answer:

```{r}
2 * pi
```

```{r}
# Gauss Hermite Quadrature
# Showing that the nodes are zeros of Hermite polynomial
Hk <- as.function(mpoly::hermite(3, kind = "he"))
nn <- mvQuad::getNodes(gg)
round(Hk(nn))
